Grounding the interaction: anchoring situated discourse in everyday human-robot interaction

### 2011-07-18 paper_knowledge.tex
Removed unused packages
Polished the introduction - moved some part elsewhere
- moved explanation on different models in oro to oro section
- moved discussion on layred architectures to the conclusion

\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{fancyvrb}
\usepackage[usenames]{color}
\usepackage{url}
\usepackage{hyperref}

@@

\usepackage{eso-pic} %% Background
%\usepackage{floatflt}

@@

@@

\usepackage{subfigure}
%diagrams
\usepackage{tikz}
%\usetikzlibrary{shapes,decorations,trees}
\usetikzlibrary{shapes,trees}

% rotate text in tabular mode
\usepackage{rotating}

@@

\begin{abstract}
This paper presents how extraction, representation and use of symbolic
knowledge from real-world perception and human-robot verbal and non-verbal
interaction can actually enable a grounded and shared model of the world that
is suitable for later high-level tasks such as dialogue understanding. We show how
the anchoring process itself relies on the situated nature of human-robot
interactions. We present an integrated approach, including a specialized symbolic
knowledge representation approach based on Description Logics, and case studies
on several robotic platforms that demonstrate these cognitive capabilities.
\end{abstract}

@@

A messy table, covered with cardboard boxes, books, video tapes... Thomas is
moving out and is packing everything with the help of Jido, his robot
(Fig.~\ref{fig|vpt}). ``Jido, give me that'', says Thomas while looking at a box that contains a
video tape. The robot smoothly grasps the tape, and hands it to the him.

@@

Working in a situated context, we need to \emph{resolve} these
semantics atoms, \ie ground them in the sensory-motor space of the robot. For
instance, ``\textit{this}'' is a demonstrative pronoun that refers in this
context to the object the human is focusing on (whatever \textit{focusing}
means). In this example Thomas is looking at something, which is a possible
cue. But it could as well be pointing at something or making reference to some
previously mentioned concept.

The next step is to extract and understand the \emph{intent} of the utterance
as thought by the agent. In our example, Thomas obviously wants an action to be performed by the robot. The action parametrization is conveyed by the semantics attached to the words and the grammatical structure of the sentence. In our example, the type of action is
given by the verb ``\textit{give}''. Assuming the robot has some procedural
knowledge attached to this symbol, the action type can be considered as
grounded for the robot. We can as well understand that the recipient of the
action is the human, the performer is the robot itself, and the object acted
upon is the tape. These are the basic \emph{thematic roles}~\cite{Gruber1965}
that can be extracted from the sentence that allow for fully grounding the action.\footnote{This analysis has been inspired on the work of Austin et al.~\cite{Austin1962}, where this type of sentences correspond to \emph{speech acts}, comprising of \emph{locutionary act} (the meaning), \emph{illocutionary} (the intent) and possibly \emph{perlocutionary acts} (implicit speaker's expectation).}

@@

In this work we propose an approach aiming at grounding the interaction with users, verbal and non-verbal, following the steps described above. To this end, we have developed three distinct, inter-related cognitive components, which we believe fulfill the minimum requisites for situated human interaction. Figure~\ref{fig|overall-process} illustrates the general context, while
Figure~\ref{fig|main-components} summarizes the interactions between these three
main components.

@@

\item \emph{Physical environment modeling} and \emph{spatial reasoning}
(grouped under the term \emph{situation assessment}): these components are in
charge of building and maintaining a coherent model of the physical world. This
model is realistic in the sense that it relies on accurate 3D models of both
manipulable objects and humans. It also has dedicated mechanisms to manage
disappearing or occluded objects.  The geometric model is used to compute
several spatial properties of the scene that actually convert the original
sensory data into symbolic beliefs, including relative locations of objects,
visibility state, gestures (such as pointing), etc.  Assuming that other agents
are also represented in the model, the same computations are applied to analyze
the scene from each agent's point of view (\ie from their \emph{perspectives}).

\item \emph{Knowledge representation and management}: the robot is endowed with
an active knowledge base that provides a logically sound symbolic model of its
beliefs of the world, as well as models for each cognitive agent the robot
interacts with. Each of these models is independent and logically consistent,
enabling reasoning on different perspectives of the world that would otherwise
be considered inconsistent (for instance, an object can be visible for the
robot but not for the human. This object can have at the same time the property
\concept{isVisible \textbf{true}} and \concept{isVisible \textbf{false}} using
two different models).  Our platform relies on OWL-DL\footnote{\emph{Web
Ontology Language - Description Logics}, a decidable subset of the first-order
logics, \url{http://www.w3.org/TR/owl2-primer/}} ontologies and features
continuous storage, querying and event triggering over the pool of facts known
by the robot.

Used in combination with the situation assessment framework, the robot is thus
able to maintain different symbolic models of the world, one per agent. This
proves an essential feature (\cite{Roy2005,Kruijff2010}) to enable
perspective-aware grounding of natural language as we will see in next
sections.

\item \emph{Dialogue input processing}: a subsystem which includes natural
language parsing capabilities, disambiguation routines and interactive concept
anchoring. We focused our efforts on three classes of utterances commonly found
in human-robot interaction: \emph{statements} (\ie new facts the human wants to
inform the robot), \emph{orders} (or more generically \emph{desires}) and
\emph{questions on declarative knowledge} (whose answers do not require
explicit planning). This would roughly cover the \emph{representative}
(sometimes referred as \emph{assertive}) and \emph{directive} illocutionary
acts in Searle's~\cite{Searle1976} classification.

@@

Based on these three components we have developed several grounding mechanisms
for situated verbal interaction and integrated them into a broader control
architecture for indoor service robots.

@@

Our work builds on top of years of work in the artificial intelligence
community around the symbol grounding issue: we can mention
Searle~\cite{Searle1980} who introduces it with the \emph{Chinese room}
metaphor, Harnard~\cite{Harnad1990} who coins the following (slightly
paraphrased in the robotic context) definition of symbol grounding:
\emph{``make the meanings of the meaningless symbol tokens intrinsic to the
robotic system"}, Coradeschi and Saffioti~\cite{Coradeschi2003} who focus on
robotics and define the meaning of \emph{anchoring} as \emph{``the process of
creating and maintaining the correspondence between symbols and sensor data
that refer to the same physical object''}, Ziemke~\cite{Ziemke1999} who
elaborates on the need of embodiment for symbol grounding, and
Sloman~\cite{Sloman2007} who criticizes the \emph{``concept empricism"} that
underlies symbol grounding.

Our contribution relates to two narrower fields: natural language in embodied
interaction contexts and knowledge building and representation in robotic
systems.

@@

Daoutis et al.~\cite{Daoutis2009} introduce one of the first complete
architectures for grounded human-robot interaction. They successfully bind
low-level percepts (including view-point independent SIFT based object
recognition) to a high-level knowledge representation and reasoning system via
a dedicated anchoring module. They base their knowledge model directly on the
\textit{ResearchCyc} ontology (including the \textit{MicroTheories} concept),
used in combination with the {\sc CycL} language. This enables second-order
logic modeling and access to a large common-sense knowledge base.

@@

Besides proposing a new integration model for sensor data, natural language and
symbolic knowledge repositories, our work extends these previous contributions
by tackling more realistic human-robot interactions: open speech; complex,
dynamic, partially unknown human environments; and fully embodied (with arms,
head, ability to move,...) autonomous robots that can perform manipulation.

Four specific contributions are presented in this work: first, we introduce a
novel and versatile knowledge base that models in a formal framework, based on
first-order logics, not only the robot's own beliefs but also every other
cognitive agent the robot interacts with.  This explicit modeling of other
agents' belief states is used for the interaction and eases the implementation
of various advanced cognitive behaviors like learning new concepts or
interactive object discrimination.

@@

Third, the same symbolic knowledge base enables richer language capabilities
for the robot.  We propose a new approach to natural language grounding that is
robust, situated and more generic than what can be found in previous work
on situated language grounding. We present several examples that include
generic recognition and semantic validation of thematic roles (\ie the
semantic roles played by the different nominal groups around a verb), which
can be translated into input for a symbolic task planner, or disambiguation
based on attention foci.

Lastly, our architecture moves away from standard layered approaches. The way
our different components interact together is not easily represented as a
layered architecture because interactions among components are mostly
bidirectional.  This is especially visible for the dialogue input processing:
this component does not simply act as an alternative perceptual input to the
symbolic database, but also actively queries previously acquired knowledge to
disambiguate and validate the newly created symbolic knowledge. This behaviour
leads us to the idea of \emph{knowledge-oriented architecture}, which is
detailed later on.

@@

In the next section we present the {\sc ORO}
(\emph{OpenRobots Ontology} server) knowledge base. We present it first, along with its objectives,
since it is the knowledge \textit{hub} of the system, used pervasively by other
components.
Section~\ref{spark} presents {\sc SPARK} (for \emph{SPAtial Reasoning \& Knowledge}), a
module that merges perceptual information with a coherent geometric model and
builds a symbolic interpretation of the world from the robot's point of view,
as well as an individual symbolic model for each agent currently present in the
environment. 
Section~\ref{dialog} covers in detail the dialogue processing component and its 
relation with ORO.
Section~\ref{experiments} presents three use-cases that were conducted on two 
different robotic platforms: the \emph{Naming} experiment, where a robot anchors new
knowledge in its model through verbal interaction; the \emph{Spy Game}, where
either the user or the robot tries to guess which object the other player is
thinking of; and a more complex experiment situated in an everyday setup, where
the robot builds models for several agents and interacts with the users using
this knowledge. 
Finally, Section~\ref{conclusion} concludes and discusses the work presented 
in this paper.

@@

@@

In this paper, we have studied knowledge flows between three components: {\it(1)}
{\sc ORO}, an ontology-based knowledge server that stores and maintains
classified RDF statements produced by other modules in agent-specific
models and allows information to be easily retrieved, either through queries or
via an event system; {\it(2)} {\sc SPARK}, the grounded, human-aware 3D model
of the environment that performs all the spatial reasoning within our
architecture, including reasoning involving motion planing (to compute
reachability of object) and perspective taking, and {\it(3)} {\sc Dialogs}, a
natural language processor that performs simple grammatical parsing of English
language, grounds the semantic content of the utterance (if necessary, also
interacts with the user to disambiguate), and eventually generates a RDF
representation of the sentence.

@@

\begin{itemize}
\item{Knowledge is explicitly stored in one central and consistent repository
of facts, accessible by all modules.}
\item{Knowledge is represented in a strict formalism (RDF statements) and
with a clearly defined vocabulary (stated in the {\tt commonsense.oro.owl}
ontology).}
\item{The first two points enable both a loosely-coupled
architecture where modules can very easily be removed or replaced by other ones
as long as they share the same semantics (modules are defined by the knowledge
they produce),}
\item{and a \emph{symbolic} reactive, event-driven approach
to supervision. By managing events at the same level as
the reasoner, we take full advantage of the inference abilities of ORO to
trigger events whose \textit{True} conditions can be inferred.}
\item{Finally, this architecture allows for the combination of very different knowledge
modalities in a single homogeneous environment, bringing mutual benefits to
components. For instance, the dialogue processing module can perfectly run
without any geometric perception, but its disambiguation routines can
transparently benefit from it when available (since richer symbolic
descriptions of objects are then available).}
\end{itemize}

>>>>

\usepackage[utf8]{inputenc}
\usepackage[usenames]{color}
\usepackage{url}
\usepackage{hyperref}

@@

@@

\usepackage{subfigure}

@@

%diagrams
\usepackage{tikz}
\usetikzlibrary{shapes,trees}

@@

\begin{abstract}
This paper presents how extraction, representation and use of symbolic
knowledge from real-world perception and human-robot verbal and non-verbal
interaction can actually enable a grounded and shared model of the world that
is suitable for later high-level tasks such as dialogue understanding. We show how
the anchoring process itself relies on the situated nature of human-robot
interactions. We present an integrated approach, including a specialized symbolic
knowledge representation system based on Description Logics, and case studies
on several robotic platforms that demonstrate these cognitive capabilities.
\end{abstract}

@@

A messy table, covered with cardboard boxes, books, video tapes... Thomas is
moving out and is packing everything with the help of Jido, his robot
(Fig.~\ref{fig|vpt}). ``Jido, give me that'', says Thomas while looking at a box that contains a
video tape. The robot smoothly grasps the tape, and hands it to him.

@@

Working in a situated context, we need to \emph{resolve} these semantics atoms,
\ie ground them in the sensory-motor space of the robot. For instance,
``\textit{that}'' is a demonstrative pronoun that refers in this context to the
object the human is focusing on.

The next step is to extract and understand the \emph{intent} of the utterance
as thought by the agent. In our example, Thomas obviously wants an action to be
performed by the robot. The action parametrization is conveyed by the semantics
attached to the words and the grammatical structure of the sentence. In our
example, the type of action is given by the verb ``\textit{give}''. Assuming
the robot has some procedural knowledge attached to this symbol, the action
type can be considered as grounded for the robot. We can as well understand
that the recipient of the action is the human, the performer is the robot
itself, and the object acted upon is the tape. The recipient, performer and object are three of the
\emph{thematic roles}~\cite{Gruber1965} that qualify the \emph{give} action. They are necessary to fully ground the sentence\footnote{This analysis has been
inspired on the work of Austin et al.~\cite{Austin1962}, where this type of
sentences correspond to \emph{speech acts}, comprising of \emph{locutionary
act} (the meaning), \emph{illocutionary} (the intent) and possibly
\emph{perlocutionary acts} (implicit speaker's expectation).}.

@@

In this work we propose an approach aiming at grounding the interaction with
users. Figure~\ref{fig|overall-process} illustrates the general contex. We have
developed three distinct, inter-related cognitive components, as illustrated in
Figure~\ref{fig|main-components}.

@@

\item \emph{Physical environment modeling} and \emph{spatial reasoning}
(grouped under the term \emph{situation assessment}): this component is in
charge of building and maintaining a coherent model of the physical world. This
model is realistic in the sense that it relies on accurate 3D models of both
manipulable objects and humans. It also has dedicated mechanisms to manage
disappearing or occluded objects.  The geometric model is used to compute
several spatial properties of the scene that actually convert the original
sensory data into symbolic beliefs, including relative locations of objects,
visibility state, gestures (such as pointing), etc.  Assuming that other agents
are also represented in the model, the same computations are applied to analyze
the scene from each agent's point of view (\ie from their \emph{perspectives}).

\item \emph{Knowledge representation and management}: the robot is endowed with
an active knowledge base that provides a symbolic model of its beliefs of the
world, as well as models for each cognitive agent the robot interacts with.
Used in combination with the situation assessment framework, this proves an
essential feature (\cite{Roy2005,Kruijff2010}) to enable perspective-aware
grounding of natural language.

Our platform relies on OWL-DL\footnote{\emph{Web Ontology Language -
Description Logics}, a decidable subset of the first-order logics,
\url{http://www.w3.org/TR/owl2-primer/}} ontologies and features continuous
storage, querying and event triggering over the pool of facts known by the
robot.

\item \emph{Dialogue processing}: the third component includes natural
language parsing capabilities, interactive disambiguation routines and concept
anchoring. We focused our efforts on three classes of utterances commonly found
in human-robot interaction: \emph{statements} (\ie new facts the human wants to
inform the robot), \emph{orders} (or more generically \emph{desires}) and
\emph{questions on declarative knowledge} (whose answers do not require
planning or prediction)\footnote{This would roughly cover the \emph{representative}
(sometimes referred as \emph{assertive}) and \emph{directive} illocutionary
acts in Searle's~\cite{Searle1976} classification.}.

@@

@@

Our work builds on top of years of work in the artificial intelligence
community around the symbol grounding issue: we can mention
Searle~\cite{Searle1980} who introduces it with the \emph{Chinese room}
metaphor, Harnard~\cite{Harnad1990} who coins the following (slightly
paraphrased in the robotic context) definition of symbol grounding:
\emph{``make the meanings of the meaningless symbol tokens intrinsic to the
robotic system"}, Coradeschi and Saffioti~\cite{Coradeschi2003} who focus on
robotics and define the meaning of \emph{anchoring} as \emph{``the process of
creating and maintaining the correspondence between symbols and sensor data
that refer to the same physical object''} and Ziemke~\cite{Ziemke1999} who
elaborates on the need of embodiment for symbol grounding.

Our contribution relates to two narrower fields: natural language in embodied
interaction contexts and knowledge acquisition and representation in robotic
systems.

@@

Daoutis et al.~\cite{Daoutis2009} introduce one of the first complete
architectures for grounded human-robot interaction. They successfully bind
low-level percepts (including view-point independent SIFT based object
recognition) to a high-level knowledge representation and reasoning system.
They base their knowledge model directly on the \textit{ResearchCyc} ontology
(including the \textit{MicroTheories} concept), used in combination with the
{\sc CycL} language. This enables second-order logic modeling and access to a
large common-sense knowledge base.

@@

Besides proposing a new integration model for sensor data, natural language and
symbolic knowledge repositories, our work extends these previous contributions
by tackling more realistic human-robot interactions: open speech; complex,
dynamic, partially unknown human environments; and fully embodied (with arms,
head, ability to move,...) autonomous robots that manipulate a large range of
household objects.

Three specific contributions are presented in this work: first, we introduce a
versatile and light-weighted knowledge base that models in a formal framework,
based on first-order logics, not only the robot's own beliefs but also every
other cognitive agent the robot interacts with.  This explicit modeling of
other agents' belief states is used for the interaction and eases the
implementation of various advanced cognitive behaviors like
False-Beliefs~\cite{Leslie2000} or interactive object discrimination.

@@

Third, the same symbolic knowledge base enables richer language capabilities
for the robot.  We propose a new approach to natural language grounding that is
robust, situated and more generic than what can be found in previous work
on situated language grounding. We present several examples that include
recognition and semantic validation of thematic roles or disambiguation
based on attention foci.

Communcation between these components is build as streams of symbolic facts,
where knowledge that the robot manipulate is made explicit.
This leads us to the idea of a \emph{knowledge-oriented architecture} which is
discussed at the end of the article.

@@

In the next section we present the {\sc ORO}
(\emph{OpenRobots Ontology} server) knowledge base. We present it first, along with its objectives,
since it is the knowledge \textit{hub} of the system, used pervasively by other
components.
Section~\ref{spark} presents {\sc SPARK} (for \emph{SPAtial Reasoning \& Knowledge}), the
component that merges perceptual information with a coherent geometric model and
builds a symbolic interpretation of the world from the robot's point of view,
as well as an individual symbolic model for each agent currently present in the
environment. 
Section~\ref{dialog} covers in detail the dialogue processing component and its 
relation with ORO.
Section~\ref{experiments} presents three use-cases that were conducted on two 
different robotic platforms: the \emph{Naming} experiment, where a robot anchors new
knowledge in its model through verbal interaction; the \emph{Spy Game}, where
either the user or the robot tries to guess which object the other player is
thinking of; and a more complex experiment situated in an everyday setup, where
the robot builds models for several agents and interacts with the users using
this knowledge. 
Finally, Section~\ref{conclusion} concludes and discusses the work presented 
in this paper.

@@

Each of these models is independent and logically consistent,
enabling reasoning on different perspectives of the world that would otherwise
be considered as globally inconsistent (for instance, an object can be visible for the
robot but not for the human. This object can have at the same time the property
\concept{isVisible \textbf{true}} and \concept{isVisible \textbf{false}} in
two different models).

@@

In this paper, we have studied knowledge streams between three components: {\it(1)}
{\sc ORO}, an ontology-based knowledge server that stores and maintains
classified RDF statements produced by other modules in agent-specific
models and allows information to be easily retrieved, either through queries or
via an event system; {\it(2)} {\sc SPARK}, the grounded, human-aware 3D model
of the environment that performs all the spatial reasoning within our
architecture, including reasoning involving motion planing (to compute
reachability of object) and perspective taking, and {\it(3)} {\sc Dialogs}, a
natural language processor that performs simple grammatical parsing of English
language, grounds the semantic content of the utterance (if necessary, also
interacts with the user to disambiguate), and eventually generates a RDF
representation of the sentence.

@@

\begin{itemize}
\item{Knowledge is explicitly stored in one central and consistent repository
of facts, accessible by all modules.}
\item{Knowledge is represented in a strict formalism (OWL statements) and
with a clearly defined vocabulary (stated in the {\tt commonsense.oro.owl}
ontology).}
\item{The first two points enable both a loosely-coupled
architecture where modules can very easily be removed or replaced by other ones
as long as they share the same semantics (modules are defined by the knowledge
they produce),}
\item{and a \emph{symbolic} reactive, event-driven approach
to supervision. By managing events at the same level as
the reasoner, we take full advantage of the inference abilities of ORO to
trigger events whose \texttt{true} conditions can be inferred.}
\item{Finally, this architecture allows for the combination of very different knowledge
modalities in a single homogeneous environment, bringing mutual benefits to
components. For instance, the dialogue processing module can perfectly run
without any geometric perception, but its disambiguation routines can
transparently benefit from it when available (since richer symbolic
descriptions of objects are then available).}
\end{itemize}

This architecture moves away from standard layered approaches: Interactions
between components are mostly bidirectional and we do not have a concept of
layers of abstraction (we do, however, have access to the robot lower levels
to execute actions, but all cognition-related modules live at the same level).
This is especially visible for the dialogue input processing: this component
does not simply act as an alternative perceptual input to the symbolic
database, but also actively queries previously acquired knowledge to
disambiguate and validate the newly created symbolic knowledge.
